---
title: "Day 2: Heatmaps and small multiples"
output: html_document
knit: (function(input, ...) {
    rmarkdown::render(
      input,
      output_dir = "C:/KU/Teaching/DataVizCourse/dev/html"
    )
  })
---

# Potentially useful libraries and starter code
```{r, message = FALSE, warning = FALSE}
library(corrplot)
library(dplyr)
library(GGally) # to make pair plots
library(ggplot2)
library(pheatmap) # a cool heatmap library
library(RColorBrewer)
library(tidyr)
library(tidyverse)

# European Quality of Life Dataset
qol <- read.csv("../data/eqls_2007and2011_with_labels.csv")
# Cardiovascular data
cardio_train <- read.csv("../data/cardio_train.csv", sep=";")
```

# Exercise 1: Exploratory data analysis on a cardiovascular training dataset
Here is a brief summary of the dataset:

 - age: in days
 - height: in cm
 - weight: in kg
 - gender (should be referred as sex): 1 = woman, 2 = man
 - cholesterol: 1 = normal, 2 = above normal, 3 = well above normal
 - glucose: 1 = normal, 2 = above normal, 3 = well above normal
 - alcohol intake: 0 = does not drink, 1 = drinks (self-reported)
 - smoking: 0 = does not smoke cigarettes, 1 = smokes (self-reported)
 - physical activity: 0 = < 1 hour of physical activity per day, 1 = >= 1 hours (self-reported)
 - cardio: 0 = no cardiovascular disease, 1 = diagnosed with CV disease

1. Print all the column names in the dataset

```{r, message = FALSE, warning = FALSE}
# Print the column headers
column_headers <- colnames(cardio_train)
print(column_headers)
```


2. Add another column called age_yrs which has the ages in years (float values)

```{r, message = FALSE, warning = FALSE}
# Convert age from days to years
cardio_train$age_yrs <- cardio_train$age / 365.25

head(cardio_train)
```

4. Do you see any null values? Are there any zeros in the age_yrs, gender*, height, weight, sys_bp, di_bp columns?

```{r, message = FALSE, warning = FALSE}
# Filter NA values
cardio_train <- na.omit(cardio_train)

# Remove rows with abnormal zeros
cardio_train <- cardio_train %>%
  filter(age_yrs != 0, gender != 0, height != 0, weight != 0, sys_bp != 0, di_bp != 0)
```

5. Examine carefully the columns sys_bp, di_bp, weight, height, and age_yrs, cholesterol. 
```{r, message = FALSE, warning = FALSE}
#     Make a plot for each distribution.
# Create a histogram for the 'weight' column
ggplot(cardio_train, aes(x = weight)) +
  geom_histogram(binwidth = 5, fill = "blue", color = "black", alpha = 0.7) +
  labs(title = "Distribution of Weight",
       x = "Weight",
       y = "Frequency")


# Create a histogram for the 'height' column
ggplot(cardio_train, aes(x = height)) +
  geom_histogram(binwidth = 5, fill = "blue", color = "black", alpha = 0.7) +
  labs(title = "Distribution of Height",
       x = "Height",
       y = "Frequency")

# Create a histogram for the 'sys_bp' column
ggplot(cardio_train, aes(x = sys_bp)) +
  geom_histogram(binwidth = 5, fill = "blue", color = "black", alpha = 0.7) +
  labs(title = "Distribution of sys_bp",
       x = "sys_bp",
       y = "Frequency")

# Create a histogram for the 'di_bp' column
ggplot(cardio_train, aes(x = di_bp)) +
  geom_histogram(binwidth = 5, fill = "blue", color = "black", alpha = 0.7) +
  labs(title = "Distribution of di_bp",
       x = "di_bp",
       y = "Frequency")      

# Create a histogram for the 'age_yrs' column
ggplot(cardio_train, aes(x = age_yrs)) +
  geom_histogram(binwidth = 5, fill = "blue", color = "black", alpha = 0.7) +
  labs(title = "Distribution of age_yrs",
       x = "age_yrs",
       y = "Frequency")      

# Create a histogram for the 'cholesterol' column
ggplot(cardio_train, aes(x = cholesterol)) +
  geom_histogram(binwidth = 5, fill = "blue", color = "black", alpha = 0.7) +
  labs(title = "Distribution of cholesterol",
       x = "cholesterol",
       y = "Frequency")
```


6. Do you see any obvious outliers in these columns? Remove them. 

```{r, message = FALSE, warning = FALSE}
# Create a boxplot for the 'height' column
ggplot(cardio_train, aes(x = factor(1), y = height)) +
  geom_boxplot(fill = "skyblue", outlier.shape = NA) +
  labs(title = "Boxplot of Height", x = NULL, y = "Height") +
  theme_minimal() +
  theme(axis.text.x = element_blank())

cardio_train <- cardio_train[cardio_train$age >= 15, , drop = FALSE]
cardio_train <- cardio_train[cardio_train$weight >= 35, , drop = FALSE]
cardio_train <- cardio_train[cardio_train$height >= 100, , drop = FALSE]
cardio_train <- cardio_train[cardio_train$sys_bp <= 240, , drop = FALSE]
cardio_train <- cardio_train[cardio_train$sys_bp >= 40, , drop = FALSE]
cardio_train <- cardio_train[cardio_train$di_bp <= 200, , drop = FALSE]
cardio_train <- cardio_train[cardio_train$di_bp >= 40, , drop = FALSE]

# Create a boxplot for the 'height' column
ggplot(cardio_train, aes(x = factor(1), y = height)) +
  geom_boxplot(fill = "skyblue", outlier.shape = NA) +
  labs(title = "Boxplot of Height", x = NULL, y = "Height") +
  theme_minimal() +
  theme(axis.text.x = element_blank())

```

7. Explore the relationships between age, height, weight, cholesterol, and glucose levels.Which plots are useful? Which are not?

```{r, message = FALSE, warning = FALSE}
# Select the columns you want to include in the pair plot
subset_data <- cardio_train[, c("age_yrs", "sys_bp", "di_bp", "cholesterol")]

# Create the pair plot
pairs(subset_data, main = "Pair Plot of age_yrs, sys_bp, di_bp, cholesterol, gluc")

# With ggplot2
points <- ggplot(cardio_train, aes(x = gender, y = sys_bp, color = factor(cholesterol))) +
  geom_point() +
  labs(title = "Pair Plot",
       x = "Sex",
       y = "Systolic Blood Pressure",
       color = "Cholesterol") +
  theme_minimal()
points
ggplot(cardio_train, aes(x = age_yrs, y = sys_bp, color = factor(cholesterol))) +
  geom_point() +
  labs(title = "Pair Plot",
       x = "Age",
       y = "Systolic Blood Pressure",
       color = "Cholesterol") +
  theme_minimal()

# With ggplot2
ggplot(cardio_train, aes(x = gender, y = weight, color = factor(cholesterol))) +
  geom_point() +
  labs(title = "Pair Plot",
       x = "Sex",
       y = "Weight",
       color = "Cholesterol") +
  theme_minimal()

ggplot(cardio_train, aes(x = age_yrs, y = weight, color = factor(cholesterol))) +
  geom_point() +
  labs(title = "Pair Plot",
       x = "Age",
       y = "Weight",
       color = "Glucose") +
  theme_minimal()
```

8. Explore the relationships between sex , cholesterol, glucose levels, smoking, alcohol, lifesyle and cardiovascular disease.

```{r, message = FALSE, warning = FALSE}
# Reshape the data into long format
cardio_train_long <- gather(cardio_train, key = "variable", value = "value", cholesterol, gluc, smoke, alco, active)

# Create the plot using ggplot2
ggplot(cardio_train_long, aes(x = variable, fill = factor(value))) +
  geom_bar(position = "dodge") +
  facet_grid(.~cardio) +
  labs(title = "Count of Categorical Variables by Cardio Status",
       x = "Variable",
       y = "Count") +
  scale_fill_manual(values = c("0" = "blue", "1" = "red")) +
  theme_minimal()  
```

9. Explore the relationships between cholesterol, glucose, smoking, alcohol and cardiovascular disease individually.
```{r, message = FALSE, warning = FALSE}
subset_data <- cardio_train[, c('cholesterol', 'gluc', 'smoke', 'alco', 'cardio')]
corr_matrix <- cor(cardio_train)
corrplot(corr_matrix, method = "color", type = "upper", order = "hclust", tl.cex = 0.7, col = colorRampPalette(c("white", "blue"))(50))

# Plot correlation heatmap
data_melted <- reshape2::melt(corr_matrix)
ggplot(data_melted, aes(Var1, Var2, fill = value)) +
  geom_tile() +
  scale_fill_gradient(low = "white", high = "blue") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))
```

10. Bonus: Violin plot
```{r, message = FALSE, warning = FALSE}
# Create a violin plot for height distribution across gender
ggplot(cardio_train, aes(x = factor(gender), y = height, fill = factor(gender))) +
  geom_violin(trim = FALSE) +
  scale_fill_manual(values = c("1" = "skyblue", "2" = "salmon")) +
  labs(title = "Violin Plot of Height Distribution by Gender",
       x = "Gender (1: Male, 2: Female)",
       y = "Height") +
  theme_minimal() +
  theme(legend.position = "top")
```

# Exercise 2: small multiples
In this exercise we will analyse mental wellbeing in the different European countries.

The index has range 0-100.

To do so, we will look at differences in distribution between countries.

Let's start with a simple plot with one density distribution plot for each country.

```{r, message = FALSE, warning = FALSE}
# Density distribution plot
ggplot(data = qol, aes(x = who5_mental_wellbeing_index, group = country_name, color = country_name)) +
  geom_density()
```

You will notice that it is not so easy to perceive differences between countries from this plot.

To solve that issue, let's do small multiples for each country!

You might need to put in extra work to adjust each panel, e.g., by putting a different colour for each panel

```{r, message = FALSE, warning = FALSE}
# Extend palette to more colours
palette <- brewer.pal(8L, "Dark2")
big_palette <- colorRampPalette(palette)(length(unique(qol$country_name)))

ggplot(data = qol, aes(x = who5_mental_wellbeing_index, color = country_name)) +
  geom_density() + # Density
  facet_wrap(~country_name) +
  theme_classic() +
  theme(
    panel.grid.major = element_line(color = "gray", linetype = "dotted"), # Grid
    axis.text.x = element_text(size = 10L), # Font size x axis
    axis.text.y = element_text(size = 10L), # Font size y axis
    legend.position = "topleft", # Legend position
    strip.background = element_blank(), # Remove borders from title
    panel.margin = unit(0.33, "cm") # Increase margin between subplots
  ) +
  labs(
    x = "Mental wellbeing index [0-100]", # Rename x label
    y = "Density" # Rename y label
  ) +
  scale_color_manual(values = big_palette) # Add color bar and palette
```


# Exercise 3: heatmaps
In this exercise we will analyse whether there are different patterns in Internet usage.

Select all the columns referring to the Internet (i.e., containing the keyword "internet"), and plot a heatmap to visualise potential differences between countries.

Save the final figure as pcardio_train. You can now use it for various reports or papers (e.g., using Overleaf!).

```{r, message = FALSE, warning = FALSE}
# Select columns with "internet" or starting with "country_name"
selected_columns <- qol %>%
  select(matches("internet|^country_name"))

# Drop rows with missing values and calculate the mean by country_name
summary <- selected_columns %>%
  drop_na() %>%
  group_by(country_name) %>%
  summarise(across(everything(), mean))

# Rename columns
summary <- summary %>%
  rename_with(
    ~ gsub("phone_internet_contact_with_", "", .), # Remove "phone_internet_contact_with" from column names
    everything()
  ) %>%
  rename(
    # rename the column "how_frequently_use_the_internet_other_than_for_work"
    other_than_work = how_frequently_use_the_internet_other_than_for_work
  )

head(summary)

# Matrix without the first column (country_name)
heatmap_matrix <- as.matrix(summary[, -1L])

# Define row and column names
rownames(heatmap_matrix) <- summary$country_name
colnames(heatmap_matrix) <- colnames(heatmap_matrix)

# Order rows alphabetically
ordered_rows <- rownames(heatmap_matrix)[order(rownames(heatmap_matrix))]

# Reorder the rows of the heatmap_matrix
heatmap_matrix_ordered <- heatmap_matrix[rev(ordered_rows), ]

head(heatmap_matrix_ordered)

# Number of colours (high value so we get something approximately continuous)
n_colors <- 100L

# Max and min of the color bar
vmin <- 1L
vmax <- 6L

# Heatmap saved as pcardio_train
pheatmap(
  heatmap_matrix_ordered,
  angle_col = 45L, # Angle for column ticks
  treeheight_row = 0L, # Remove dendrogram
  treeheight_col = 0L, # Remove dendrogram
  col = colorRampPalette(c("#c13639", "white", "#2f79b5"))(n_colors),
  breaks = seq(vmin, vmax, length.out = n_colors + 1L),
  filename = "../img/heatmap_day2.pdf" # Save as Pcardio_train
)
```


