---
title: 'Exercises: Prinicpal Component Analysis (PCA)'
output:
  html_document: default
  pdf_document: default
date: "2023-09-26"
---


Welcome to the exercise session on Principle Component Analysis. In this class
we will go through the steps of conducting PCA on a dataset relating to kidney 
failure. 

We will be using a couple of additional packages that are required to conduct PCA 
in R. The first is called 'corrr' and is used for running a PCA, and the second is 
'corrplot', which plots the output of corrr.


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(corrr)
library(corrplot)
library(dplyr)
library(factoextra)
library(forcats)
library(ggplot2)
```

First, read in the data and have a look at the columns that are available. 

```{r read data}
kidney_data <- read.csv("../data/chronic_kidney_disease_full.csv")
```

Now we need to check whether there are NAs in our data - remember that this can 
break our analysis if we don't check for it first! 

```{r check NAs}
colSums(is.na(kidney_data))

# is.na checks for NA values in each cell of the 
# dataframe. colSums returns the sum of the columns - if any one entry is NA, 
# then the colSums will give us an NA! So this is a neat way of checking for the
# presence of NA values. 
```
Check which of the variables are categorical and which are continuous - remember
that we need continuous data for PCA! Extract the columns that correspond to continuous
data. How many variables do you get?

```{r}
head(kidney_data)

# Select the conitnuous variables
data_for_PCA <- kidney_data %>% select(age:sg, bgr:rbcc)

# Remove NA values, and apply character-to-numeric conversion
data_for_PCA <- data_for_PCA  %>%
  mutate_if(is.character, as.numeric) %>% na.omit()

# Scale (subtract mean and divide by SD) and convert to dataframe
data_for_PCA <- as.data.frame(scale(data_for_PCA))

head(data_for_PCA)
```
Now plot a correlation matrix with your standardised data. For a very simple 
correlation matrix you can use the cor function from the corrr package and then
use corrplot, but you could also try creating a correlation matrix in ggplot. 


```{r}
# Correlation matrix
corr_matrix <- cor(data_for_PCA)

# Correlation matrix plot
corrplot(
  corr_matrix,
  # Only plot the lower triangular
  type = 'lower', 
  # Visualisation method: show ellipses
  # eccentricity scaled to the correlation value
  method = 'ellipse',
  # Sort by angular order of the eigenvectors
  order = 'AOE',
  # Rotate the text labels
  tl.srt = 45
)
```

Now, we can perform the PCA. We can use the command prcomp to perform the PCA on the 
scaled data. 


```{r}
# Perform PCA
PCA <- prcomp(corr_matrix)

# SD of the principal components
sdevs <- PCA$sdev

# Total variance
total_var <- sum(sdevs^2)

# Get proportion variance (divide SD^2 by total variance)
proportion_of_variance <- data.frame(Proportion_of_Variance = (sdevs^2) / total_var * 100)

proportion_of_variance$Component <- as.character(row.names(proportion_of_variance))


# Scree plot
scree_plot <- ggplot(
  proportion_of_variance, aes(
    fct_reorder(Component, -Proportion_of_Variance), Proportion_of_Variance, group = 1)
  ) +
  # Add line plot
  geom_line(color = 'dodgerblue') +
  # Add scatter plot
  geom_point(color = 'dodgerblue') +
  # Change y limit, xy labels, title, and theme
  ylim(0, 100) + 
  xlab('Component') + 
  ylab('Percentage of Variance') + 
  ggtitle('Percentage of total variance explained by each component') + 
  theme_bw()

scree_plot
```


We can also use fviz_pca_var to plot the each variable in terms of the principal components. 


```{r}
fviz_pca_var(PCA, col.var = 'black')
```


Finally, we can use a cos2 plot to transform the principal components back into our original:

```{r}
fviz_cos2(PCA, choice = "var", axes = 1)
```

We can also use PCA to predict whether or not a patient will have chronic kidney disease. 
We split the data into a train set, on which we perform the PCA, and a test set, for which we try to predict whether each patient has chronic kidney disease. 

We can then plot the principal components for each of these patients on a scatter plot, 
and see whether the patients are separated by their kidney condition. 

```{r}
# Take the continuous data 
data_for_prediction <- kidney_data %>% select(X, age:sg, bgr:rbcc)

# Remove NAs
data_for_PCA <- data_for_prediction %>%
  mutate_if(is.character, as.numeric) %>% na.omit()

# Add the category labels, i.e. does each person have chronic kidney disease
data_for_PCA$class <- kidney_data %>% filter(X %in% data_for_PCA$X) %>%
  select(class)

# Split the data into testing and training - choose a random sample from the full 
# dataset
train <- data_for_PCA %>% sample_frac(0.70)
test  <- dplyr::anti_join(data_for_PCA, train, by = 'X')

#Scale the training set and produce a correlation matrix
scale_train <- as.data.frame(scale(train %>% select(age:sg, bgr:rbcc)))
corr_matrix <- cor(scale_train)
corrplot(corr_matrix, type = 'lower', method = 'ellipse', order = 'AOE', 
         tl.srt = 45) 

# Perform the PCA on the scaled training set 
PCA <- prcomp(scale_train, scale. = TRUE, center = TRUE)

# Scale and perform PCA on the test set
scale_test <- as.data.frame(scale(test %>% select(age:sg, bgr:rbcc)))
pred <- predict(PCA, newdata = scale_test)
train_df <- as.data.frame(PCA$x[, 1:2])
train_df$Diagnosis <- unname(unlist(train$class))

# Transform the testing data using the principal components. 
predict_df <- as.data.frame(pred[, 1:2])
predict_df$Diagnosis <- unname(unlist(test$class))

# Plot the results
scatter <- ggplot(predict_df, aes(x = PC1, y = PC2, color = Diagnosis)) +
  geom_point() + theme_bw() 
scatter
```




