---
title: "Day 3: Networks and maps"
output: html_document
---

# Potentially useful libraries and starter code
```{r, message = FALSE, warning = FALSE}
# install.packages("devtools")
# devtools::install_github("ropensci/rnaturalearthhires")

library(ggplot2) # general purpose plotting
library(ggspatial) # scale bars and north arrows
library(rnaturalearth) # map data
library(rnaturalearthdata)# map data
# library(rnaturalearthhires)# map data 
library(sf) #'simple features' package
library(tmap) # interactive map
```

# Execise 1: plot the world

Use the rnaturalearth and ggplot2 libraries to get the country outlines on a world map

```{r, message = FALSE, warning = FALSE, name = "Ex1"}
# We use the rnaturalearth library to get the countries
world_rne <- ne_countries(
  scale="medium", # map scale
  returnclass = "sf" # return a Simple Features dataframe
)

# We use geom_sf from ggplot to draw simple boundaries and outlines
ggplot(data = world_rne) +
  geom_sf() # geom to visualise sf objects
```

# Exercise 2: plot countries and continents

Do 4 different plots for the maps of:
 * 1. Kenya
 * 2. Denmark
 * 3. Europe
 * 4. Asia

```{r, message = FALSE, warning = FALSE, name = "Ex2"}
# Use ne_countries to get natural earth world country polygons
kenya <- ne_countries(
  country="kenya", # country name
  type="countries", # country type
  scale='large',
  returnclass = "sf"
)

# Plot Kenya
kenya_map <- ggplot(data = kenya) +
  geom_sf()

kenya_map
```
 
# Exercise 3: adding details

For the Kenya map, add:
  * the scale
  * the orientation (north/south)
  * coordinates (latitude/longitude)

```{r, message = FALSE, warning = FALSE, name = "Ex3"}
kenya_map_detailed <- kenya_map +
  # Add labels
  labs(
    x = "Longitude",
    y = "Latitude",
    title = "Kenya"
  ) +
  # Add the scale on the bottom left
  annotation_scale(location = "bl", width_hint = 0.4) +
  # Add a north arrow on the bottom left (bl)
  annotation_north_arrow(
    location = "bl", 
    which_north = "true", 
    pad_x = unit(0.0, "in"), 
    pad_y = unit(0.2, "in"),
    style = north_arrow_fancy_orienteering
)

kenya_map_detailed
```
  
# Exercise 4: even more details

Augment your Kenya plot by adding a point for the capital, Nairobi, and 3 other major cities: Mumbase, Nakuru, Ruiru.
Do the same for Denmark as well.
You might have to do a bit of googling to get the coordinates :)

```{r, message = FALSE, warning = FALSE, name = "Ex4"}
# Kenya data for major cities
kenya_city_data <- data.frame(
  city_name=c("Nairobi", "Mumbasa", "Nakuru", "Ruiru"),
  lat=c(-1.2921,-4.0435,-0.3031,-1.1483),
  lon=c(36.8219,39.6682,36.0800,36.9605)
)

kenya_map_detailed_with_cities <- kenya_map_detailed +
  # Add points for each city
  geom_point(
    data = kenya_city_data, 
    mapping = aes(x = lon, y = lat), 
    colour = "red"
  ) +
  # Add annotation text for each city
  geom_text(
    data = kenya_city_data, 
    mapping=aes(x=lon, y=lat, label=city_name), 
    nudge_y = 0.5, 
    color="darkblue"
  )

kenya_map_detailed_with_cities
```

# Exercise 5: Thematic maps

*Warning!* You might have to delete your current environment before doing this exercise and/or do it on a separate R script.

World is a data.frame with a special column that contains a geometry for each row, in this case polygons.
```{r, message = FALSE, warning = FALSE, name = "Ex5"}
# Interactive mode
tmap_mode("view")

# Load World Data
data("World")

# Show the first rows
head("World")
```

In order to plot it in tmap, you first need to specify it with tm_shape. Layers can be added with the + operator, in this case tm_polygons.

```{r, message = FALSE, warning = FALSE, name = "Ex5p2"}
# Create a tmap element with tm_shape
# Draw polygons for the population estimation with tm_polygons
tm_shape(World) + tm_polygons("pop_est")
```

 
# Exercise 6: adding more layers

Load data about rivers, land, and major metropolitan areas.

Start with a plot of all land in terms of elevation, then add country borders and finally the locations of the major metropolitan areas.

```{r, message = FALSE, warning = FALSE, name = "Ex5p2"}
# Load data for metros, rivers, and land
data(World, metro, rivers, land)

# Print the first rows

# Metro has a geometry of style POINT
print(head(metro))

# Rivers has a geometry of style LINESTRING
print(head(rivers))

# Land contains spatial data of type "stars", it contains raster data (hint: column refsys)
print(land)
```

```{r, message = FALSE, warning = FALSE, name = "Ex5p2"}
# Plot the elevation from the land data
tm_shape(land) +
  tm_raster("elevation", palette = terrain.colors(10)) +
  # Now plot the borders from the World data
  tm_shape(World) +
  tm_borders("white", lwd = .5) +
  # Plot the country names with 3 letters (iso_a3)
  tm_text("iso_a3", size = "AREA") +
  # Plot the metropolotian areas
  tm_shape(metro) +
  # Add points, whose size depend on pop2020
  tm_symbols(col = "red", size = "pop2020", scale = .5) +
  tm_legend(show = FALSE)
```

# Exercise 7: this time for Africa

Let's dive back into Africa. Plot the life expectancy in African countries

```{r, message = FALSE, warning = FALSE}
# Select Africa
africa <- World[World$continent == "Africa",]

tm_shape(africa) +
  # Plot life expectancy
  tm_polygons(col = "life_exp", palette = "BuGn") +
  # Add title
  tm_layout(title = "Life Expectancy in Africa")
```

# Exercise 8: your turn!

Have a look at the columns in the ```africa``` dataset. Try to plot a few variables individually on the map.

You can also try to use ```tm_facets``` to plot several variables simultaneously


```{r, message = FALSE, warning = FALSE}
# Select Africa
africa <- World[World$continent == "Africa",]

tm_shape(africa) +
  # Plot economy
  tm_polygons(col = "gdp_cap_est", palette = "BuGn")
  # Add title
  tm_layout(title = "GDP per capita of African countries")

# You can plot several columns at once with tm_facets
tm_shape(africa) +
  # Plot economy
  tm_polygons(col = c("footprint", "gdp_cap_est"), palette = "BuGn") +
  tm_facets(sync = TRUE, ncol = 2)
```
