---
title: "Day 4: Interactive visualisation"
output: html_document
knit: (function(input, ...) {
    rmarkdown::render(
      input,
      output_dir = "C:/KU/Teaching/DataVizCourse/dev/html"
    )
  })
---

# Potentially useful libraries and starter code
```{r, message = FALSE, warning = FALSE}
library(dplyr)
library(ggplot2)
library(gridExtra)
library(Hmisc)
library(plotly)

# European quality of life dataset
qol <- read.csv("../data/eqls_2007and2011_with_labels.csv")

# DK Energy consumption
dk_cons <- read.csv("../data/dk_energy_consumption_2021.csv")

head(dk_cons)
```

# Exercise 0: ggplot to plotly

Take your plot from Day 1/Exercise 3 and make it interactive with plotly!
```{r}
summary_work_hours_first_job <- qol %>%
  select(how_many_hours_work_per_week_in_1st_job, country_name) %>%
  group_by(country_name) %>%
  summarise(data = list(smean.cl.boot(cur_data(), conf.int = .95, B = 1000L, na.rm = TRUE))) %>%
  tidyr::unnest_wider(data)


fig <- ggplot(summary_work_hours_first_job, aes(x = Mean, y = reorder(country_name, Mean))) +
  # geom_bar(stat = "identity", aes(fill = Mean)) +
  geom_point(aes(color = Mean), size = 3L) +
  geom_errorbar(aes(xmin = Lower, xmax = Upper, y = country_name)) +
  scale_fill_distiller(type = "seq", palette = "Blues", direction = 1) +
  theme_classic() +
  theme(
    axis.text.x = element_text(size = 11L),
    axis.text.y = element_text(size = 11L)
  ) +
  labs(
    x = "Work hours (Mean +/- 95% CI)",
    y = "",
  )

plotly_fig <- ggplotly(fig)

# Display the interactive plot
plotly_fig
```

# Exercise 1: Sankey diagram
Do a Sankey diagram to represent the sources and targets of energy consumption in Denmark in 2021.

To do so:
 * Filter out the entries where the output energy is 0.
 * Build a data frame of nodes mapped to an integer index.
 * Build a data frame of links where:
    * Source = integer node from the previous dataframe (the energy source)
    * Target = integer node from the previous dataframe (the target sector)
    * Value = value of energy consumption for that link
```{r, fig.height=8}
nodes <- data.frame(name = unique(c(unique(dk_cons$Source), unique(dk_cons$Sector))))

links <- data.frame(
  source = match(dk_cons$Source, nodes$name) - 1,
  target = match(dk_cons$Sector, nodes$name) - 1,
  value = dk_cons$TJ
)

links <- subset(links, value != 0)

fig <- plot_ly(
  type = "sankey",
  orientation = "h",
  node = list(
    pad = 15,
    thickness = 20,
    line = list(color = "black", width = 0.5),
    label = unique(c(dk_cons$Source, dk_cons$Sector))
  ),
  link = list(
    source = links$source,
    target = links$target,
    value = links$value
  )
)

fig <- fig %>% layout(
  title = "DK Energy Consumption (2021; in TJ)",
  font = list(
    size = 14
  )
)

fig
```

# Exercise 2: building your first Shiny app
cf. app.R in ```day4```

Build an interactive app to analyse how people in Europe perceive and trust their institutions (using the same old European quality of life dataset).

We already got some work ready for you by:
 1. Calculating the ```trust``` column, which tells how much on average the responders _trust_ the different public institutions (e.g., police, parliament...)
 2. Calculating the ```quality``` column, which tells how much on average the responders _rate_ the different public institutions (e.g., police, parliament...)
 3. Coding the UI side! We put a row with a sidebar, and only one input: a country (which can be selected from the user).

Your task is to plot a scatterplot for any input country with:
 * x = trust
 * y = quality
 * Two x- and two y- intercepts: one for the national averages in red, one for the sample averages in blue (from the dataset)

Pay attention to the _reactive_ parts of your plot, and use plotly for interactive visualisation :)
