---
title: "Day 2: Heatmaps and small multiples"
output: html_document
---

# Potentially useful libraries and starter code
```{r, message = FALSE, warning = FALSE}
library(dplyr)
library(ggplot2)
library(pheatmap)
library(RColorBrewer)
library(tidyr)

# European Quality of Life Dataset
qol <- read.csv("../data/eqls_2007and2011_with_labels.csv")
# Cardiovascular data
cardio_train <- ("../data/cardio_train.csv")
```

# Exercise 1: Exploratory data analysis on a cardiovascular training dataset
Here is a brief summary of the dataset:

 - age: in days
 - height: in cm
 - weight: in kg
 - gender (should be referred as sex): 1 = woman, 2 = man
 - cholesterol: 1 = normal, 2 = above normal, 3 = well above normal
 - glucose: 1 = normal, 2 = above normal, 3 = well above normal
 - alcohol intake: 0 = does not drink, 1 = drinks (self-reported)
 - smoking: 0 = does not smoke cigarettes, 1 = smokes (self-reported)
 - physical activity: 0 = < 1 hour of physical activity per day, 1 = >= 1 hours (self-reported)
 - cardio: 0 = no cardiovascular disease, 1 = diagnosed with CV disease

1. Print all the column names in the dataset

2. The dataset should contain a column "age" representing age in days. Add a column "age_yrs" which contains the age in years (as a float/decimal value)

3. Now that you have the data, what would be the first step you would do? Write in brief your thinking of methodology.

4. Do you see any null values? Are there any zeros in the age_yrs, gender*, height, weight, sys_bp, di_dp columns? If you find any, remove those rows.

5. Examine carefully the columns sys_bp, di_dp, weight, height, and age_yrs, cholesterol. Plot the distribution of each column in a separate plot. What do you see? Which distributions are continuous? Which ones are discrete? Do you examine any skewness in the distributions? 

6. Do you see any obvious outliers in these columns? Remove them. Which plot was the most useful to remove outliers?

7. Explore the relationships between age, height, weight, cholesterol, and glucose levels. Which plots are useful? Which are not?

8. Explore the relationships between sex, cholesterol, glucose levels, smoking, alcohol, lifesyle and cardiovascular disease. 

9. Explore the relationships between cholesterol, glucose, smoking, alcohol and cardiovascular disease individually.

10. Bonus: a violin plot showing the distribution of any variable with respect to sex


```{r, message = FALSE, warning = FALSE}
# Your code here!
```

# Exercise 2: small multiples
In this exercise we will analyse mental wellbeing in the different European countries.

The index has range 0-100.

To do so, we will look at differences in distribution between countries.

Let's start with a simple plot with one density distribution plot for each country.

```{r, message = FALSE, warning = FALSE}
# Your code here!
```

You will notice that it is not so easy to perceive differences between countries from this plot.

To solve that issue, let's do small multiples for each country!

You might need to put in extra work to adjust each panel, e.g., by putting a different colour for each panel

```{r, message = FALSE, warning = FALSE}
# Your code here!
```


# Exercise 3: heatmaps

In this exercise we will analyse whether there are different patterns in Internet usage.

Select all the columns referring to the Internet (i.e., containing the keyword "internet"), and plot a heatmap to visualise potential differences between countries.

Save the final figure as pdf. You can now use it for various reports or papers (e.g., using Overleaf!).
```{r, message = FALSE, warning = FALSE}
# Your code here!
```


