---
title: 'Exercises: Prinicpal Component Analysis (PCA)'
output:
  html_document: default
  pdf_document: default
date: "2023-09-21"
---


Welcome to the exercise session on Principle Component Analysis. In this class
we will go through the steps of conducting PCA on a dataset relating to kidney 
failure. 

We will be using a couple of additional packages that are required to conduct PCA 
in R. The first is called 'corrr' and is used for running a PCA, and the second is 
'corrplot', which plots the output of corrr.


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(corrr)
library(corrplot)

library(dplyr)
library(ggplot2)
```

First, read in the data and have a look at the columns that are available. 

```{r read data}
# Read kidney disease data here
```

Now we need to check whether there are NAs in our data - remember that this can 
break our analysis if we don't check for it first! 

```{r check NAs}
  # ////////////////////////////////////////////////////////////
  #                      YOUR CODE HERE
  # ////////////////////////////////////////////////////////////
```
Check which of the variables are categorical and which are continuous - remember
that we need continuous data for PCA! Extract the columns that correspond to continuous
data. How many variables do you get?

```{r}
  # ////////////////////////////////////////////////////////////
  #                      YOUR CODE HERE
  # ////////////////////////////////////////////////////////////
  
  
```
Now plot a correlation matrix with your standardised data. For a very simple 
correlation matrix you can use the cor function from the corrr package and then
use corrplot, but you could also try creating a correlation matrix in ggplot. 


```{r}
  # ////////////////////////////////////////////////////////////
  #                      YOUR CODE HERE
  # ////////////////////////////////////////////////////////////

```
Now, we can perform the PCA. We can use the command prcomp to perform the PCA on the 
scaled data. 


```{r}
  # ////////////////////////////////////////////////////////////
  #                      YOUR CODE HERE
  # ////////////////////////////////////////////////////////////
```



We can also use fviz_pca_var to plot the each variable in terms of the principal components. 


```{r}
library(factoextra)
fviz_pca_var(PCA, col.var = 'black')

```


Finally, we can use a cos2 plot to transform the principal components back into our original  
```{r}
fviz_cos2(PCA, choice = "var", axes = 1)
```
We can also use PCA to predict whether or not a patient will have chronic kidney disease. 
We split the data into a train set, on which we perform the PCA, and a test set, for which we try to predict whether each patient has chronic kidney disease. 

We can then plot the principal components for each of these patients on a scatter plot, 
and see whether the patients are separated by their kidney condition. 

```{r}
# Take the continuous data 
data_for_prediction <- kidney_data %>% select(X, age:sg, bgr:rbcc)
# Remove NAs
data_for_PCA <- data_for_prediction %>%
  mutate_if(is.character, as.numeric) %>% na.omit()
# Add the category labels, i.e. does each person have chronic kidney disease
data_for_PCA$class <- kidney_data %>% filter(X %in% data_for_PCA$X) %>%
  select(class)

# Split the data into testing and training - choose a random sample from the full 
# dataset
train <- data_for_PCA %>% sample_frac(0.70)
test  <- dplyr::anti_join(data_for_PCA, train, by = 'X')

#Scale the training set and produce a correlation matrix

scale_train <- as.data.frame(scale(train %>% select(age:sg, bgr:rbcc)))
corr_matrix <- cor(scale_train)
corrplot(corr_matrix, type = 'lower', method = 'ellipse', order = 'AOE', 
         tl.srt = 45) 

# Perform the PCA on the scaled training set 
PCA <- prcomp(scale_train, scale. = TRUE, center = TRUE)


scale_test <- as.data.frame(scale(test %>% select(age:sg, bgr:rbcc)))
pred <- predict(PCA, newdata = scale_test)
train_df <- as.data.frame(PCA$x[, 1:2])
train_df$Diagnosis <- unname(unlist(train$class))

# Transform the testing data using the principal components. 
predict_df <- as.data.frame(pred[, 1:2])
predict_df$Diagnosis <- unname(unlist(test$class))
# Plot the results
scatter <- ggplot(predict_df, aes(x = PC1, y = PC2, color = Diagnosis)) +
  geom_point() + theme_bw() 
scatter
```




