---
title: "Day 3: Networks and maps"
output: html_document
---

# Potentially useful libraries and starter code
```{r, message = FALSE, warning = FALSE}
# install.packages("devtools")
# devtools::install_github("ropensci/rnaturalearthhires")

library(ggplot2) # general purpose plotting
library(ggspatial) # scale bars and north arrows
library(rnaturalearth) # map data
library(rnaturalearthdata)# map data
# library(rnaturalearthhires)# map data 
library(sf) #'simple features' package
library(tmap) # interactive map
```

# Execise 1: plot the world

Use the rnaturalearth and ggplot2 libraries to get the country outlines on a world map

```{r, message = FALSE, warning = FALSE}
# ////////////////////////////////////////////////////////////
#                      YOUR CODE HERE
# ///////////////////////////////////////////////////////////
```

# Exercise 2: plot countries and continents

Do 4 different plots for the maps of:
 * 1. Kenya
 * 2. Denmark
 * 3. Europe
 * 4. Asia

```{r, message = FALSE, warning = FALSE}
# ////////////////////////////////////////////////////////////
#                      YOUR CODE HERE
# ///////////////////////////////////////////////////////////
```
 
 # Exercise 3: adding details
 For the Kenya map, add:
  * the scale
  * the orientation (north/south)
  * coordinates (latitude/longitude)
 Do the same for Denmark as well.
```{r, message = FALSE, warning = FALSE}
# ////////////////////////////////////////////////////////////
#                      YOUR CODE HERE
# ///////////////////////////////////////////////////////////
```
  
# Exercise 4: even more details

Augment your Kenya plot by adding a point for the capital, Nairobi, and 3 other major cities: Mumbase, Nakuru, Ruiru.

You might have to do a bit of googling to get the coordinates :)

```{r, message = FALSE, warning = FALSE}
# ////////////////////////////////////////////////////////////
#                      YOUR CODE HERE
# ///////////////////////////////////////////////////////////
```

# Exercise 5: Thematic maps

*Warning!* You might have to delete your current environment before doing this exercise and/or do it on a separate R script.


Let's first load the "World" data and visualise the first lines
```{r, message = FALSE, warning = FALSE}
# ////////////////////////////////////////////////////////////
#                      YOUR CODE HERE
# ///////////////////////////////////////////////////////////
```

Now, plot the population of each country on a world map. Make it interactive!

```{r, message = FALSE, warning = FALSE}
# ////////////////////////////////////////////////////////////
#                      YOUR CODE HERE
# ///////////////////////////////////////////////////////////
```

# Exercise 6: adding more layers

Load data about rivers, land, and major metropolitan areas.

Start with a plot of all land in terms of elevation, then add country borders and finally the locations of the major metropolitan areas.

```{r, message = FALSE, warning = FALSE}
# ////////////////////////////////////////////////////////////
#                      YOUR CODE HERE
# ///////////////////////////////////////////////////////////
```

# Exercise 7: this time for Africa

Let's dive back into Africa. Plot the life expectancy in African countries

```{r, message = FALSE, warning = FALSE}
# ////////////////////////////////////////////////////////////
#                      YOUR CODE HERE
# ///////////////////////////////////////////////////////////
```

# Exercise 8: your turn!

Have a look at the columns in the ```africa``` dataset. Try to plot a few variables individually on the map.

You can also try to use ```tm_facets``` to plot several variables simultaneously

```{r, message = FALSE, warning = FALSE}
# ////////////////////////////////////////////////////////////
#                      YOUR CODE HERE
# ///////////////////////////////////////////////////////////
```
